---
title: 'Assignment 5: Data Visualization'
author: "Alonso Velasquez"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options:
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 Loading Packages
library(tidyverse)
library(here)
library(cowplot)
library(ggthemes)
here()
#2 Reading in datasets
PeterPaul.chem.nutrients <- read.csv(
  here('./Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv'),
  stringsAsFactors = TRUE)
NIWO_litter <- read.csv(
  here('./Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv'),
  stringsAsFactors = TRUE)

#Fixing dates
PeterPaul.chem.nutrients$sampledate <- ymd(PeterPaul.chem.nutrients$sampledate)
NIWO_litter$collectDate <- ymd(NIWO_litter$collectDate)
class(PeterPaul.chem.nutrients$sampledate)
class(NIWO_litter$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
#Create a base plot to play with
my_theme <- theme_classic(base_size = 14, base_family = "serif") +
  theme(plot.title = element_text(color = '#4F4F4F'),
        axis.ticks = element_line(color = '#c77975'),
        axis.line = element_line(color = '#F0F0F0'),
        axis.title.x = element_text(color = '#4F4F4F'),
        axis.title.y = element_text(color = '#4F4F4F'),
        axis.text = element_text(color = '#c77975'),
        legend.text = element_text(color = '#4F4F4F'),
        legend.title = element_text(color = '#4F4F4F'),
        legend.background = element_rect(fill = '#F0F0F0'),
        plot.background = element_rect(fill = '#F0F0F0'),
        panel.background = element_rect(fill = '#c77975'),
        panel.grid.major = element_line(color = "#F0F0F0"),  # main grid lines
        panel.grid.minor = element_line(color = "#F0F0F0"),
        rect = element_rect(color = "#767676"))

# Theme building layout used
# my_theme <- theme_base()  + 
 # theme(
    #line =              element_line(),
    #rect =              element_rect(),
    #text =              element_text(),
    
    # Modified inheritance structure of text element
    #plot.title =        element_text(),
    #axis.title.x =      element_blank(),
    #axis.title.y =      element_blank(),
    #axis.text =         element_text(),
    
    # Modified inheritance structure of line element
    #axis.ticks =        element_line(),
    #panel.grid.major =  element_line(),
    #panel.grid.minor =  element_blank(),
    
    # Modified inheritance structure of rect element
    #plot.background =   element_rect(),
    #panel.background =  element_rect(),
    #legend.key =        element_rect(),
    
    # Modifiying legend.position
    #legend.position = 'right',
    
    #complete = TRUE
   # )

#Sample plot to play around with theme

PeterPaul.chem.nutrients %>% 
  filter(depth == 0 & year4 == 1999) %>% 
  ggplot(
    mapping = aes(
      x=sampledate,
      y=temperature_C,
      color=lakename)
    ) + 
  geom_line() +
  my_theme

# setting my_theme to be the default theme

set_theme(my_theme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
lake_colors <- c(
  "Peter Lake" = "#84665D",
  "Paul Lake"  = "#F6E5E3")

PeterPaul.chem.nutrients %>% 
  ggplot(mapping = aes(
      x=tp_ug,
      y=po4,
      color=lakename)) +
    xlim(0, 150) +
    ylim(0, 50) +
    geom_point(size = 1.3, alpha = 0.8) +
    geom_smooth(method = lm) +
    scale_color_manual(values = lake_colors) + 
    labs(color = "Lake Name") +
    xlab("Total Phosphorus (µg/L)") +
    ylab("PO4 (µg/L)")
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 
#identifying class of months
class(PeterPaul.chem.nutrients$month)
#ensuring all 12 are here
unique(PeterPaul.chem.nutrients$month)

#building the plot (a) with factors
p_a <- PeterPaul.chem.nutrients %>%
  ggplot(aes(x = factor(month,levels=1:12,labels=month.abb), y = temperature_C, color = lakename)) +
  geom_boxplot() +
  my_theme +
  #scale_x_discrete(name='Month', drop=FALSE) +
  scale_color_manual(values = lake_colors) +
  ylab("Temperature C°") +
  xlab("Month") +
  labs(color = "Lake Name", title = "Temperature Boxplot by Lake per Month")  

print(p_a)

#building the plot (b) with factors
p_b <- PeterPaul.chem.nutrients %>%
  ggplot(aes(x = factor(month,levels=1:12,labels=month.abb), y = tp_ug, color = lakename)) +
  geom_boxplot() +
  my_theme +
  #scale_x_discrete(name='Month', drop=FALSE) +
  scale_color_manual(values = lake_colors) +
  ylab("Total Phosphorus (µg/L)") +
  xlab("Month") +
  labs(color = "Lake Name", title = "Total Phosphorus Boxplot by Lake per Month") + 
  theme(panel.background = element_rect(fill = '#9cc2c3'),
        axis.ticks = element_line(color = '#9cc2c3'),
        axis.text = element_text(color = '#9cc2c3'))

print(p_b)

#building the plot (c) with factors
p_c <- PeterPaul.chem.nutrients %>%
  ggplot(aes(x = factor(month,levels=1:12,labels=month.abb), y = tn_ug, color = lakename)) +
  geom_boxplot() +
  my_theme +
  #scale_x_discrete(name='Month', drop=FALSE) +
  scale_color_manual(values = lake_colors) +
  ylab("Total Nitrogen (µg/L)") +
  xlab("Month") +
  labs(color = "Lake Name", title = "Total Nitrogen Boxplot by Lake per Month") + 
  theme(panel.background = element_rect(fill = '#91a792'),
        axis.ticks = element_line(color = '#91a792'),
        axis.text = element_text(color = '#91a792'))

print(p_c)

#combining them into a single plot using cowplot
p_a2 <- p_a +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none",
        axis.title.y = element_text(size = 8)) +
  labs(title = "Monthly Variance in Nutrients at Peter and Paul Lakes")

p_b2 <- p_b +
  labs(title = NULL) +  
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 8))

# Remove title only from p3
p_c2 <- p_c + 
  labs(title = NULL) +
  theme(legend.position = "none",
        axis.title.y = element_text(size=8, margin = margin(t = -5, r = 5, b = 10, l = 0)))

plot_grid(p_a2, p_b2, p_c2, 
          ncol = 1,   
          align = "v") 
        
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Generally, concentrations of TP and TN appear to rise and spike minutely during the summer months, peaking around July and August. They start to fall again during September. Temperature, of course rises during the summer months. Although TP and TN could be related to temperature, they could also be related to water stratification or unrelated events that occur during the summer. More analysis is needed to understand this phenomenon.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6 Litter Subplot separated by color

litter_colors <- c(
  "evergreenForest" = "#6a743f",
  "grasslandHerbaceous"  = "#32909a",
  "shrubScrub" = "#e6d995")

NIWO_litter %>% 
  filter(functionalGroup == 'Needles') %>% 
  ggplot(mapping = aes(x=collectDate, y=dryMass, color=nlcdClass)) +
  geom_point() +
  xlab("Collection Date") +
  ylab("Dry Mass (g)") +
  labs(color = 'NLCD Class',title = "Needles NIWO Litter Data Over Time by NLCD Class") +
  scale_x_date(date_breaks = "4 months", date_labels = "%b%Y") +
  scale_color_manual(values = litter_colors) +
  theme(axis.text.x = (element_text(size=6)))

#7 Facets
NIWO_litter %>% 
  filter(functionalGroup == 'Needles') %>% 
  ggplot(mapping = aes(x=collectDate, y=dryMass)) +
  geom_point(size = 0.5) +
  xlab("Collection Date") +
  ylab("Dry Mass (g)") +
 # scale_x_date(date_breaks = "4 months", date_labels = "%b%Y") + #removing scale breaks with faceting because it is more clunky with three rows
  theme(axis.text.x = (element_text(size=7))) +
  facet_wrap(vars(nlcdClass), nrow = 3)
```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: Plot 7 is more effective from a data visualization perspective, because you are visualizing comparisons more easily. With different colors on the same plot, the shape of the data is less easily visible because there are other colored points on the same plot. When faceting for so many points, it's easier to see if there is a different curve or shape to the data. However, some may prefer being able to see data compared on the same plot. If the data were visualized as lines, it may be more helpful, but for point data, I believe seeing the shape of the points on different facets is more effective.