---
title: "Assignment 8: Time Series Analysis"
author: "Alonso Velasquez"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_TimeSeries.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

```{r}
library(tidyverse)
library(lubridate)
#install.packages("trend")
library(trend)
#install.packages("zoo")
library(zoo)
#install.packages("Kendall")
library(Kendall)
#install.packages("tseries")
library(tseries)

#checking working directory
library(here)

#setting ggplot theme
my_theme <- theme_classic(base_size = 14, base_family = "serif") +
  theme(plot.title = element_text(color = '#4F4F4F'),
        axis.ticks = element_line(color = '#c77975'),
        axis.line = element_line(color = 'white'),
        axis.title.x = element_text(color = '#4F4F4F'),
        axis.title.y = element_text(color = '#4F4F4F'),
        axis.text = element_text(color = '#c77975'),
        legend.text = element_text(color = '#4F4F4F'),
        legend.title = element_text(color = '#4F4F4F'),
        legend.background = element_rect(fill = 'mistyrose'),
        plot.background = element_rect(fill = 'white'),
        panel.background = element_rect(fill = '#c77975'),
        panel.grid.major = element_line(color = "mistyrose"),  
        panel.grid.minor = element_line(color = "mistyrose"),
        rect = element_rect(color = "white"))

set_theme(my_theme)
```


2. Import the ten datasets from the Ozone_TimeSeries folder in the Raw data folder. These contain ozone concentrations at Garinger High School in North Carolina from 2010-2019 (the EPA air database only allows downloads for one year at a time). Import these either individually or in bulk and then combine them into a single dataframe named `GaringerOzone` of 3589 observation and 20 variables. 

```{r, message = FALSE}
#1: reading in datasets
EPAair_O3_GaringerNC2010_raw <- read.csv(here('./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2010_raw.csv'), stringsAsFactors = TRUE)
EPAair_O3_GaringerNC2011_raw <- read.csv(here('./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2011_raw.csv'), stringsAsFactors = TRUE)
EPAair_O3_GaringerNC2012_raw <- read.csv(here('./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2012_raw.csv'),stringsAsFactors = TRUE)
EPAair_O3_GaringerNC2013_raw <- read.csv(here('./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2013_raw.csv'),stringsAsFactors = TRUE)
EPAair_O3_GaringerNC2014_raw <- read.csv(here('./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2014_raw.csv'),stringsAsFactors = TRUE)
EPAair_O3_GaringerNC2015_raw <- read.csv(here('./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2015_raw.csv'),stringsAsFactors = TRUE)
EPAair_O3_GaringerNC2016_raw <- read.csv(here('./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2016_raw.csv'),stringsAsFactors = TRUE)
EPAair_O3_GaringerNC2017_raw <- read.csv(here('./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2017_raw.csv'),stringsAsFactors = TRUE)
EPAair_O3_GaringerNC2018_raw <- read.csv(here('./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2018_raw.csv'),stringsAsFactors = TRUE)
EPAair_O3_GaringerNC2019_raw <- read.csv(here('./Data/Raw/Ozone_TimeSeries/EPAair_O3_GaringerNC2019_raw.csv'),stringsAsFactors = TRUE)

#2: combining cases
EPAair_O3_GaringerNC_2010_2019_raw <- bind_rows(EPAair_O3_GaringerNC2010_raw,EPAair_O3_GaringerNC2011_raw,EPAair_O3_GaringerNC2012_raw,EPAair_O3_GaringerNC2013_raw,EPAair_O3_GaringerNC2014_raw, EPAair_O3_GaringerNC2015_raw, EPAair_O3_GaringerNC2016_raw, EPAair_O3_GaringerNC2017_raw, EPAair_O3_GaringerNC2018_raw, EPAair_O3_GaringerNC2019_raw, .id = 'dataset')

```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order of data frames within this function so that the final dimensions are 3652 rows and 3 columns. Call your combined data frame GaringerOzone. 

```{r}
# 3: Setting date column as date class
EPAair_O3_GaringerNC_2010_2019_raw$Date <- mdy(EPAair_O3_GaringerNC_2010_2019_raw$Date)
# 4: Wrangling
EPAair_03_GaringerNC_cleaned <- EPAair_O3_GaringerNC_2010_2019_raw %>%
  select(Date, Daily.Max.8.hour.Ozone.Concentration, DAILY_AQI_VALUE)
# 5: New dataframe as sequence
Days <- as.data.frame(seq.Date(from=as.Date("2010/01/01"),
                                 to=as.Date("2019/12/31"),
                                 by="day"))
colnames(Days) <- "Date"

# 6: Left joining and finding gaps in data
GaringerOzone <- left_join(Days, EPAair_03_GaringerNC_cleaned, by = "Date")

```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case, we will plot actual concentrations in ppm, not AQI values. Format your axes accordingly. Add a smoothed line showing any linear trend of your data. Does your plot suggest a trend in ozone concentration over time?

```{r, fig.cap='Daily Ozone Concentrations 2010-2019'}
#7: Plotting data
ggplot(GaringerOzone, aes(x = Date, y = Daily.Max.8.hour.Ozone.Concentration)) + 
  geom_line(col='seashell') +
  geom_smooth(method = lm, fill = 'salmon', col='seashell4', se = TRUE) +
  labs(x = "Date", y = "Ozone (ppm)")
```

>Answer: The plot shows a few different patterns. First, there is big seasonal variation in ozone throughout the year, peaking towards summer and declining towards the winter. Second, there seems to be a gradual decline in annual ozone since 2010, as the trend line shows a slight negative trend. Lastly, the geom_smooth shows little shading around the trend line, indicating its standard error is low and there is high confidence in the trend line from the linear model.

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station? 

8. Use a linear interpolation to fill in missing daily data for ozone concentration. Why didn't we use a piecewise constant or spline interpolation?

```{r}
#8: Using linear interpolation to fill in ozone concentrations
GaringerOzone$Daily.Max.8.hour.Ozone.Concentration = 
  zoo::na.approx(GaringerOzone$Daily.Max.8.hour.Ozone.Concentration)
```

> Answer: Our entire dataset is connected by linear lines, and our model suggests the trend over time is linear. There is also consistent seasonal fluctuation, so linear interpolation makes more sense over spline or piecewise constant. Although piecewise constant could be used, ozone is constantly fluctuating with season, so it makes sense to find intermediate values to fill gaps instead of the nearest neighbor values. Spline does not make sense to use here because points are connected by straight lines and our relationship appears to be linear, there is not a quadratic "curved" relationship. Therefore, linear interpolation makes the most sense for filling missing data.


9. Create a new data frame called `GaringerOzone.monthly` that contains aggregated data: mean ozone concentrations for each month. In your pipe, you will need to first add columns for year and month to form the groupings. In a separate line of code, create a new Date column with each month-year combination being set as the first day of the month (this is for graphing purposes only)

```{r}
#9: grouping by month and finding average ozone
GaringerOzone.monthly <- GaringerOzone %>% 
  mutate(Month = month(Date)) %>% 
  mutate(Year = year(Date)) %>% 
  group_by(Month, Year) %>% 
  summarize(mean_ozone = mean(Daily.Max.8.hour.Ozone.Concentration))

#Making new column for graphing purposes
GaringerOzone.monthly <- GaringerOzone.monthly %>% 
  mutate(MonthYear = as.Date(paste(Year, Month, "01", sep = "-")))
```


10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` and base it on the dataframe of daily observations. Name the second `GaringerOzone.monthly.ts` and base it on the monthly average ozone values. Be sure that each specifies the correct start and end dates and the frequency of the time series.

```{r}
#10: Creating time series objects
GaringerOzone.daily.ts <- 
  ts(GaringerOzone$Daily.Max.8.hour.Ozone.Concentration, frequency = 365, start = c(2010,1))
GaringerOzone.monthly.ts <- ts(GaringerOzone.monthly$mean_ozone, frequency = 12, start = c(2010,1))
```


11. Decompose the daily and the monthly time series objects and plot the components using the `plot()` function.

```{r}
#11: Decomposing time series objects
GaringerOzone.daily.decomposed <- stl(GaringerOzone.daily.ts, s.window = "periodic")
GaringerOzone.monthly.decomposed <- stl(GaringerOzone.monthly.ts, s.window = "periodic")

#plotting components
plot(GaringerOzone.daily.decomposed)
plot(GaringerOzone.monthly.decomposed)
```


12. Run a monotonic trend analysis for the monthly Ozone series. In this case the seasonal Mann-Kendall is most appropriate; why is this?

```{r}
#12: Running monotonic trend analysis
GaringerOzone.monthly.trend1 <- 
  Kendall::SeasonalMannKendall(GaringerOzone.monthly.ts)

summary(GaringerOzone.monthly.trend1)

#inspecting results for individual seasons
GaringerOzone.monthly.trend2 <- trend::smk.test(GaringerOzone.monthly.ts)

summary(GaringerOzone.monthly.trend2)

```

> Answer: The seasonal Mann-Kendall test is the most appropriate because ozone has a prominent seasonal component. We are interested in identifying if ozone has changed over time while accounting for that seasonal component -- Monotonic trends are gradual shifts over time, and to assess that we need to incorporate seasonal fluctuation in our test! 

13. Create a plot depicting mean monthly ozone concentrations over time, with both a geom_point and a geom_line layer. Edit your axis labels accordingly.

```{r, fig.cap='Monthly Mean Ozone Concentrations 2010-2019'}
# 13
ggplot(GaringerOzone.monthly, aes(x = MonthYear, y = mean_ozone)) + 
  geom_line(col='seashell') +
  geom_smooth(method = lm, fill = 'seashell', col='seashell4', se = TRUE) +
  geom_point(colour='seashell', fill='salmon', shape=21) +
  labs(x = "Date", y = "Ozone (ppm)", title="Monthly Mean Ozone Concentrations")
  


```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: We want to assess if we see a monotonic trend--a gradual shift over time--in our monthly average ozone concentrations over the period of 2010 to 2019. Has monthly mean ozone concentrations changed over time? Although we see a slight negative trend in our plotted data, to prove the null hypothesis that there is no monotonic trend wrong, we need to incorporate seasonal fluctuation in our test. Running a seasonal Mann-Kendall test to assess gradual change in ozone concentrations, our p-values are a bit large (with our 2-sided pvalue being =0.163). Although there seems to be a negative trend in ozone concentrations over time, we cannot rule out the null hyopthesis from this test.

15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`. Hint: Look at how we extracted the series components for the EnoDischarge on the lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series. Compare the results with the ones obtained with the Seasonal Mann Kendall on the complete series.

```{r}
#15: subtracting seasonal component from our ts object
GaringerOzone.monthly.nonseas <- (GaringerOzone.monthly.ts - GaringerOzone.monthly.decomposed$time.series[,1])

#making sure it worked
plot(GaringerOzone.monthly.decomposed)
plot(GaringerOzone.monthly.nonseas)


#16: Running Mann Kendall on non-seasonal Ozone monthly series
GaringerOzone.monthly.nonseas.trend <- trend::mk.test(GaringerOzone.monthly.nonseas)
# Inspect results
GaringerOzone.monthly.nonseas.trend

```

> Answer: After running the Mann Kendall test on the non-seasonal data, there is still no significant trend that can disprove the null hypothesis (p-value = 0.1039). Although it is a smaller p-value than our seasonal test (two-sided p-value = 0.163), it is still greater than alpha. Therefore, we cannot conclude that monthly ozone averages have gradually changed over time.
