---
title: "Assignment 3: Data Exploration"
author: "Alonso"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r setup,  message = FALSE}
#Setting up packages
#library(tidyverse)
#library(lubridate)
#library(here)
#getting wd
getwd()
here()
#uploading datasets
Neonics <- read.csv(
  file = here('./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv'),
  stringsAsFactors = TRUE,
)
Litter <- read.csv(
  file = here('./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv'),
  stringsAsFactors = TRUE,
)
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: We need to understand which species of insects affected. Are they pests or beneficial for local biodiversity? We also need to understand the environmental health impacts. What impacts could human consumption or runoff of neonicotinoids have?

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Insects have a large contribution to the decomposition of forest litter and woody debris, and are important to a health forest floor system. If local insect populations are being affected by pesticides, it's possible that this will have a downstream effect on local litter systems.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.the dry weight of litterfall and fine woody debris are collected from litter traps by plant functional type
 2. Mass data for each collection event are measured separately for the following functional groups, to an accuracy of 0.01 grams. Weights < 0.01g are reported and may indicate presence of a given functional group, identified in the sortng process, but not present at detectable masses.
 3. The finest resolution at which temporal data are reported is the daysOfTrapping, the range between setDate and collectDate.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Q5}
#Finding dimensions of the dataset

print(dim(Neonics))

#Neonics df has 4623 rows and 30 columns

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r Q6}
#summarizing the Neonics "effect" column
help(summary)
sort(summary(Neonics$Effect))

```

> Answer: Popultation, Mortality, and Behavior are the top three studied effects, with 1803, 1493, and 360 observations, respectively


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r Q7}
#Identifying the name of the common name species column, summarizing obersvations in column
head(Neonics)
sort(summary(Neonics$Species.Common.Name))
```

> Answer: Other (670), Honey Bee (667), Parasitic Wasp (285), Buff Tailed Bumblebee (183), Carniolan Honey Bee (152), Bumble Bee (140), Italian Honeybee (113) are the species with the most observations (I included other as not part of the six most commonly studied species)

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r Q8}
#Identifying class of 'Conc.1..Author'
class(Neonics$Conc.1..Author.)

```

> Answer: It's not numeric because several values have "/" or "NR" instead of numbers.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Q9}
#generating a plot with number of studies conducted by publication year
ggplot(data = Neonics, aes(x = Publication.Year)) +
  geom_bar()
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Q10}
#Same graph but now with a color aesthetic by different test locations
ggplot(data = Neonics, aes(x = Publication.Year, fill = Test.Location)) +
  geom_bar()

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common testing locations are the Lab,  although before 2005 'Field natural' was the most common 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Q11}
#Plotting a bar graph of Endpoint counts
ggplot(data = Neonics, aes(x = Endpoint)) +
  geom_bar() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```

> Answer: 'LOEL and 'NOEL' are the most common endpoints. LOEL stands for Lowest-observable-effect-level, which is referring to the endpoint where the lowest dose (concentration) producing effects  were significantly different (as reported by authors) from responses of controls. NOEL stands for No-observable-effect-level, which is referring to the endpoint where the highest dose (concentration) producing effects were not significantly different from responses of controls according to author's reported statistical test. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Q12}
#exploring the class of Litter's collectDate -- it's a factor and not a date
class(Litter$collectDate)
#converting the factor into a date
Litter$collectDate <- ymd(Litter$collectDate)
class(Litter$collectDate)
```

13. Using the `unique` function, list the different plotIDs sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Q13}
help(unique)
unique(Litter$plotID)
summary(Litter$plotID)
```

> Answer: There are 12 different plotIDs sampled at Niwot Ridge. the 'unique' function gives the number of different unique values in the column, while summary gives the count of each unique value that comes up in the column. 'unique' seems more useful when working with many unique values and wanting to know how many there are.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Q14}
#creating a bar graph of functionalGroup counts
ggplot(data = Litter, aes(x = functionalGroup)) +
  geom_bar()
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r Q15}
help(geom_boxplot)
#making boxplot
ggplot(Litter, aes(x=functionalGroup, y=dryMass)) + 
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#making violinplot
ggplot(Litter, aes(x=functionalGroup, y=dryMass)) + 
  geom_violin() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: Most of the functional groups have really tiny values, with a few larger values sprinkled inbetween. This can be well illustrated in a boxplot, but violin plots exaggerate the imbalanced distribution of values. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles and twigs/branches
